---
title: "Analyzing the horizontal orientation of the crustal stress adjacent to plate boundaries"
author: "Tobias Stephan"
date: "2023-08-06"
output:
  html_document:
    highlight: tango
    #theme: united
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: yes
subtitle: 3 - Geo-Data
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tectonicr)
library(dplyr)
library(ggplot2)
library(patchwork)
```

## Geo-Data Analysis

Angles are measured from North pole

Angles measured on different geo-locations are affected by distortion


### Synthetic dataset

```{r data, echo = TRUE}
# Synthetic dataset ####
## Generate in PoR ####
n.data <- 100
mean <- 90

#set.seed(16)
data_90_0 <- tibble(
  lat = runif(n.data, 80, 90),
  lon = runif(n.data, 0, 360),
  azi = rvm(n.data, mean = mean, kappa = 10), # sample from von Mises distribution
  prd = mean,
  x = lon,
  y = lat
) |>
  sf::st_as_sf(coords = c("x", "y"), crs = "WGS84") |>
  mutate(
    dev = abs(90 - azi),
    cdist = circular_distance(azi, prd)
  )
```


Statistics of our dataset
```{r stat_funct}
circular_summary <- function(x) {
  x |>
    sf::st_drop_geometry() |>
    summarise(
      mean = circular_mean(azi),
      median = circular_median(azi),
      sd = circular_sd(azi),
      var = circular_var(azi),
      CI95 = confidence_angle(azi),
      kappa = tectonicr:::est.kappa(azi)
    )
}
```

```{r stats, message=FALSE, echo = TRUE}
circular_summary(data_90_0)
rose(data_90_0$azi, col = "grey", main = "Center coordinate: 90°N, 0°E")

ggplot() +
  geom_spoke(data = data_90_0, aes(lon, lat, angle = deg2rad(90 - azi), color = cdist), radius = 1, position = "center_spoke") +
  coord_equal() +
  scale_color_viridis_c() +
  theme_bw()

```

```{r rotate_funct, echo = TRUE}
rotate_to_new_coordinate <- function(x0, lat0 = 90, lon0 = 0, lat1 = 0, lon1 = 0) {
  euler_por <- data.frame(lat = lat0, lon = lat0)
  euler_geo <- data.frame(lat = lat1, lon = lon1)
  northpole_por <- data.frame(lat = lat1, lon = lon1 + 180)

  sf::st_crs(x0) <- PoR_crs(euler_geo)
  x1 <- PoR_to_geographical_sf(x0, euler_geo)

  x1 <- cbind(
    x1,
    sf::st_coordinates(x1) |> as.data.frame() |> rename(lon = X, lat = Y)
  ) |>
    select(lat.1, lon.1) |>
    rename(lat = lat.1, lon = lon.1)

  x1$azi <- PoR_shmax(x0, northpole_por)
  return(x1)
}
```

```{r rotate_data, warning=FALSE, echo = TRUE}
data_0_0 <- rotate_to_new_coordinate(data_90_0)
data_22_0 <- rotate_to_new_coordinate(data_90_0, lat1 = 22.5, lon1 = 33)
data_45_0 <- rotate_to_new_coordinate(data_90_0, lat1 = 45)
data_67_0 <- rotate_to_new_coordinate(data_90_0, lat1 = 67.5, lon1 = -33)

rbind(
  circular_summary(data_90_0) |> mutate(lat = 90),
  circular_summary(data_67_0) |> mutate(lat = 67.5),
  circular_summary(data_45_0) |> mutate(lat = 5),
  circular_summary(data_22_0) |> mutate(lat = 22.5),
  circular_summary(data_0_0) |> mutate(lat = 0)
)
```

```{r p0, echo = TRUE}
ggplot() +
  geom_sf(data = eulerpole_smallcircles(data.frame(lat = 0, lon = 0), n = 90) |> filter(between(d, 0, 10)), aes(color = as.factor(d)), lty = 1) +
  geom_point(data = NULL, aes(0, 0, color = as.factor(0)), size = 7.5, show.legend = FALSE) +
  geom_spoke(data = data_0_0, aes(lon, lat, angle = deg2rad(90 - azi)), radius = 1, lwd = 1, position = "center_spoke") +
  geom_sf(data = data_0_0) +
  scale_color_viridis_d("Distance to center (°)") +
  theme_bw() +
  coord_sf(
    default_crs = sf::st_crs(4326),
    crs = sf::st_crs(
      "+proj=ortho +lat_0=0"
    )
  )
```

```{r p1, echo = TRUE}
world <- rnaturalearth::ne_countries(scale = "small", returnclass = "sf") |> sf::st_geometry()

data_plot <- ggplot() +
  geom_sf(data = world, alpha = .75, color = NA) +
  geom_sf(data = eulerpole_smallcircles(data.frame(lat = 90, lon = 0), n = 90) |> filter(between(d, 0, 10)), aes(color = "centered at 90°N"), lty = 1) +
  geom_sf(data = eulerpole_smallcircles(data.frame(lat = 67.5, lon = -33), n = 90) |> filter(between(d, 0, 10)), aes(color = "centered at 67.5°N"), lty = 1) +
  geom_sf(data = eulerpole_smallcircles(data.frame(lat = 45, lon = 0), n = 90) |> filter(between(d, 0, 10)), aes(color = "centered at 45°N"), lty = 1) +
  geom_sf(data = eulerpole_smallcircles(data.frame(lat = 22.5, lon = 33), n = 90) |> filter(between(d, 0, 10)), aes(color = "centered at 22.5°N"), lty = 1) +
  geom_sf(data = eulerpole_smallcircles(data.frame(lat = 0, lon = 0), n = 90) |> filter(between(d, 0, 10)), aes(color = "centered at 0°N"), lty = 1) +
  scico::scale_color_scico_d(NULL) +
  theme_bw()

data_plot + coord_sf(
  default_crs = sf::st_crs(4326),
  crs = sf::st_crs(
    "+proj=ortho +lat_0=0"
  )
) +
  data_plot + coord_sf(
    default_crs = sf::st_crs(4326),
    crs = sf::st_crs(
      "+proj=ortho +lat_0=46"
    )
  ) +
  data_plot + coord_sf(
    default_crs = sf::st_crs(4326),
    crs = sf::st_crs(
      # "+proj=natearth"
      "+proj=ortho +lat_0=90"
    )
  ) +
  plot_layout(guides = "collect")
```

Let's have a look at the rose diagrams for the different locations:

```{r p2, echo = TRUE}
lats <- seq(0, 90, 22.5)
cols <- scico::scico(length(lats), palette = "batlow")
names(cols) <- as.character(lats)


par(mfrow = c(2, 3))
rose(data_90_0$azi, main = "90°N", sub = paste0("Var: ", round(circular_var(data_90_0$azi), 3)), binwidth = 20, col = cols["90"])
rose(data_67_0$azi, main = "67.5°N", sub = paste0("Var: ", round(circular_var(data_67_0$azi), 3)), binwidth = 20, col = cols["67.5"])
rose(data_45_0$azi, main = "45°N", sub = paste0("Var: ", round(circular_var(data_45_0$azi), 3)), binwidth = 20, col = cols["45"])
rose(data_22_0$azi, main = "22.5°N", sub = paste0("Var: ", round(circular_var(data_22_0$azi), 3)), binwidth = 20, col = cols["22.5"])
rose(data_0_0$azi, main = "0°N", sub = paste0("Var: ", round(circular_var(data_0_0$azi), 3)), binwidth = 20, col = cols["0"])
```

Ooops, it seems like the distribution changes with latitude????


Let's have a closer look. I model the same stress field for s sequence of latitude from -90 to 90 with a smaller increment. Then I plot the distribution parameters variance and mean against the latitude:
```{r loop, warning = FALSE, echo = TRUE}
increment <- 1
stats <- circular_summary(data_90_0) |> mutate(lat = 90)
for (i in seq(increment, 180, increment)) {
 xi <-  rotate_to_new_coordinate(data_90_0, lat1 = 90 - i) |> 
    circular_summary() |> 
    mutate(lat = 90 - i)
  stats <- rbind(stats, xi)
}
```

```{r var_vs_lat, echo = TRUE}
ggplot(stats, aes(var, abs(lat))) +
  geom_point() +
  scale_y_continuous(breaks = seq(-90, 90, 22.5)) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 90), expand = F) +
  ggplot(stats, aes(mean, abs(lat), color = var)) +
  geom_point() +
  scale_y_continuous(breaks = seq(-90, 90, 22.5)) +
  coord_cartesian(ylim = c(0, 90), expand = F) +
  scale_color_viridis_c()
```

We see that there is a clear dependence of both variance and mean on the geographic latitude.
