---
title: "Programming with R --- A Beginners’ Guide for Geoscientists"
subtitle: "5 - Maps"
author: "Tobias Stephan"
date: "03/02/2022"
output: 
  html_document:
    theme: "united"
    highlight: "tango"
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
      smooth_scroll: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Packages we need:
```{r packages, echo = TRUE, message = FALSE, warning = FALSE}
gc(verbose = FALSE)
require(dplyr)
require(squish)

# pLotting
require(ggplot2)
require(scico)
require(patchwork)
require(ggrepel)

# geospatial packages
require(sp)
require(sf)
require(raster)
require(rgeos)
require(rgdal)

# spatial interpolation
# require(gstat)
# require(dismo)
require(phylin)


# my functions and stuff for the course
source("course_stuff.R")
```

## Maps

### Seismics at BC coast
```{r data}
earthquakes <- read.csv("Data/Map/earthquakes_2000-2020.csv", sep = ",") %>% arrange(depth)
head(earthquakes)
```



### Basic maps
```{r map1, echo = TRUE, message = FALSE, warning = FALSE}

map1 <- ggplot() + # extent of the map
  borders() + # political borders
  geom_point(
    data = earthquakes,
    aes(longitude, latitude, color = depth), size = 2
  ) +
  scale_color_scico(palette = "bilbao", name = "Depth (km)") # a colorblind-friendly color-palette

map1 +
  coord_map(
    projection = "mercator", # the projection
    xlim = range(earthquakes$longitude), ylim = range(earthquakes$latitude)
  ) + # extent of the map
  labs(title = "mercator", subtitle = "equally spaced straight meridians, conformal, straight compass courses") +
  map1 +
  coord_map(
    projection = "azequidistant", # the projection
    xlim = range(earthquakes$longitude), ylim = range(earthquakes$latitude)
  ) + # extent of the map
  labs(title = "azequidistant", subtitle = "equally spaced parallels, true distances from pole") +
  map1 +
  coord_map(
    projection = "azequalarea", # the projection
    xlim = range(earthquakes$longitude), ylim = range(earthquakes$latitude)
  ) + # extent of the map
  labs(title = "azequalarea", subtitle = "equal-area") +
  map1 +
  coord_map(
    projection = "albers", parameters = c(50, 58.5), # the projection
    xlim = range(earthquakes$longitude), ylim = range(earthquakes$latitude)
  ) + # extent of the map
  labs(title = "albers", subtitle = "equal-area, true scale on lat0 and lat1")
gc(verbose = FALSE)
```

```{r map2}
ggplot(data = earthquakes) +
  borders(fill = alpha("grey", .25)) + # political borders
  geom_point(aes(longitude, latitude, color = depth, size = mag), alpha = .75) +
  scale_color_scico(palette = "hawaii", name = "Depth (km)", direction = -1, breaks = seq(0, 30, 5)) +
  scale_size_binned(name = "Magnitude") +
  # scale_shape()+
  # scale_alpha()+
  coord_map(
    projection = "azequalarea", # the projection
    xlim = range(earthquakes$longitude), ylim = range(earthquakes$latitude)
  ) + # extent of the map
  labs(title = "Seismicity of western British Columbia", subtitle = "Data source: ANSS Comprehensive Earthquake Catalog", caption = "equal-area projection") +
  theme_bw()
```



### Import Shape-files

```{r read.shp, message=FALSE}
# load shape file as 'simple feature'
plates <- sf::st_read("Data/Map/plate_boundaries.shp") # plate boundaries
sf::st_crs(plates) <- "WGS84"
```

```{r shp.prepare, message=FALSE, warning=FALSE}
plates.df <- sf::st_drop_geometry(plates) %>% mutate(id = as.character(seq_along(plates$Code)))
plates.sp <- sf::as_Spatial(plates) %>% # convert into "sp" feature
  broom::tidy() %>%
  left_join(plates.df)

gc()
```

```{r map.plate}
ggplot(data = earthquakes %>% arrange(depth)) +
  geom_path(data = plates.sp, aes(long, lat, group = group), lwd = 2, color = colorblind[7]) +
  borders(fill = alpha("grey", .25)) + # political borders
  geom_point(aes(longitude, latitude, color = depth, size = mag), alpha = .75) +
  scale_color_scico(palette = "hawaii", name = "Depth (km)", direction = -1, breaks = seq(0, 30, 5)) +
  scale_size_binned(name = "Magnitude") +
  coord_map(
    projection = "azequalarea", # the projection
    xlim = range(earthquakes$longitude), ylim = range(earthquakes$latitude)
  ) + # extent of the map
  labs(title = "Seismicity of western British Columbia", subtitle = "Data source: ANSS Comprehensive Earthquake Catalog", caption = "equal-area projection") +
  theme_bw()
```
### Create a spatial object

Create a lines object
```{r line1}
line.pts <- cbind(
  lon = c(-120, -130),
  lat = c(52, 48)
)

line.df <- as.data.frame(cbind(line.pts))



line.sf <- sf::st_linestring(line.pts) %>%
  sf::st_geometry() %>%
  sf::st_set_crs("+proj=longlat +ellps=WGS84 +no_defs")
```

Plot the new object as a profile line:
```{r plot_line}
ggplot() +
  geom_path(
    data = plates.sp,
    aes(long, lat, group = group),
    lwd = 2,
    color = colorblind[7]
  ) +
  borders(fill = alpha("grey", .25)) +
  geom_point(
    data = earthquakes,
    aes(longitude, latitude, color = depth, size = mag),
    alpha = .75
  ) +
  scale_color_scico(
    palette = "hawaii",
    name = "Depth (km)",
    direction = -1,
    breaks = seq(0, 30, 5)
  ) +
  scale_size_binned(name = "Magnitude") +

  # adds the profile line
  geom_line(data = line.df, aes(lon, lat), lty = 2) +

  # end points
  geom_point(data = line.df, aes(lon, lat), size = 2) +

  # label the points
  geom_label_repel(data = line.df, aes(lon, lat), label = c("A", "B")) +
  coord_map(
    projection = "azequalarea",
    # the projection
    xlim = range(earthquakes$longitude),
    ylim = range(earthquakes$latitude)
  ) + # extent of the map
  labs(
    title = "Seismicity of western British Columbia",
    subtitle = "Data source: ANSS Comprehensive Earthquake Catalog",
    caption = "equal-area projection"
  ) +
  theme_bw()

gc(verbose = FALSE)
```

Save line into ESRI shapefile:
```{r line2, echo=FALSE, eval=FALSE, include=TRUE}
sf::st_write(line.sf, "Data/Map/my_line.shp")
```



### Import Raster

```{r raster_load}
# dem <- raster::raster('Data/DEM/srtm_13_02.tif')
dem <- raster::raster("Data/Map/ETOPO_vancouver.tif")

# first glimpse
plot(dem)
```
#### Crop a raster
```{r raster_crop}
crop_extent <- sf::st_bbox(
  c(
    xmin = min(earthquakes$longitude),
    xmax = max(earthquakes$longitude),
    ymin = min(earthquakes$latitude),
    ymax = max(earthquakes$latitude)
  ),
  crs = sf::st_crs("WGS84")
) %>%
  sf::st_as_sfc() %>%
  sf::as_Spatial()

dem.cropped <- crop(dem, crop_extent)
plot(dem.cropped)
plot(crop_extent, add = TRUE)


dem.df <- as(dem.cropped, "SpatialPixelsDataFrame") %>%
  as.data.frame()
colnames(dem.df) <- c("elevation", "lon", "lat")
```
#### Save raster
```{r write_raster, eval=FALSE, include=TRUE}
raster::writeRaster(dem.cropped, "Data/Map/dem_cropped.tif", overwrite = TRUE)
```


```{r raster_stats, warning=FALSE, message=FALSE}
# Stats numbers
dem.df$elevation %>% summary()

# Histogram
hist(dem.df$elevation)
```
#### Plot the raster

```{r plot_raster, echo=TRUE}
map2 <- ggplot() +

  # add the raster with coloring (CPU intensive!!!)
  geom_raster(data = dem.df, aes(x = lon, y = lat, fill = elevation)) +
  geom_path(
    data = plates.sp,
    aes(long, lat, group = group),
    lwd = 2,
    color = colorblind[7]
  ) +
  borders(fill = alpha("grey", .25)) +
  geom_point(
    data = earthquakes,
    aes(longitude, latitude, color = depth, size = mag),
    alpha = .75
  ) +
  geom_path(data = line.df, aes(lon, lat), lty = 2) +
  geom_point(data = line.df, aes(lon, lat), size = 2) +
  geom_label_repel(data = line.df, aes(lon, lat), label = c("A", "B")) +
  scale_color_scico(
    palette = "hawaii",
    name = "Depth (km)",
    direction = -1,
    breaks = seq(0, 30, 5)
  ) +
  scale_size_binned(name = "Magnitude") +

  # colorblind color palette for topography
  scale_fill_scico(
    palette = "oleron",
    name = "Topography (m a.s.l.)",
    # legend title
    limits = c(-max(dem.df$elevation), max(dem.df$elevation)),
    #
    oob = scales::squish,
    #
    breaks = seq(-5000, 5000, 1000)
  ) + # format legend labels
  coord_cartesian(
    xlim = range(earthquakes$longitude),
    ylim = range(earthquakes$latitude),
    expand = FALSE
  ) +
  labs(
    title = "Seismicity of western British Columbia",
    subtitle = "Data source: ANSS Comprehensive Earthquake Catalog",
    caption = "equal-area projection"
  ) +
  theme_bw()

map2
```


## Save a map
```{r save_map, echo=FALSE, eval=FALSE, include=TRUE}
ggsave(map2, filename = "Data/DEM/my_map.pdf", width = 11, height = 8.5)
```


### Swath-profile from a DEM

```{r swath, echo=TRUE, message=FALSE, warning=FALSE}
# extracts the swath profile and some statistics
# WARNING CPU-intensive
swath.profile <- swathR(
  coords = line.pts,
  raster = dem.cropped,
  crs = "+proj=longlat +ellps=WGS84 +no_defs",
  dist = .2,
  k = 1
)

swath.stats <- as.data.frame(swath.profile[[1]])
swath.data <- swath.profile[[2]]
swath.lines <- swath.profile[[3]]
```

```{r plot_swath, echo=TRUE}
swath.data.line <- data.frame(dist = seq_along(swath.data[[2]]), elevation = swath.data[[2]])

ggplot(swath.data.line, aes(dist, elevation)) +
  coord_cartesian(expand = FALSE) +
  geom_ribbon(aes(ymin = elevation - swath.stats$std.dev.[2], ymax = elevation + swath.stats$std.dev.[2]), fill = "grey") +
  geom_line(color = colorblind[4]) +
  labs(x = "Distance (degree)", y = "Elevation (m)")
```



### "Heat map" --- Spatial interpolation

reproject into equal area projection
```{r}
earthquakes.sp <- earthquakes
coordinates(earthquakes.sp) <- ~ longitude + latitude
sp::proj4string(earthquakes.sp) <-
  sp::CRS("+proj=longlat +ellps=WGS84 +no_defs")

aea <-
  sp::CRS(
    "+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000 +y_0=0 +datum=NAD83 +units=m +no_defs"
  )

earthquakes.sp.aea <- sp::spTransform(earthquakes.sp, aea)


```

initialize the interpolation grid
```{r grid}
extent <- crop_extent@bbox %>%
  t() %>%
  as.data.frame()
coordinates(extent) <- ~ x + y
proj4string(extent) <- "+proj=longlat +ellps=WGS84 +no_defs"

extent.aea <- sp::spTransform(extent, aea)

spacing <- 150000 # m
grid <-
  expand.grid(
    x = seq(
      from = extent.aea@coords[1, 1],
      to = extent.aea@coords[2, 1],
      by = spacing
    ),
    y = seq(extent.aea@coords[1, 2], to = extent.aea@coords[2, 2], by = spacing)
  ) # expand points to grid
# coordinates(grid) <- ~ x + y
# proj4string(grid) <- aea
# gridded(grid) <- TRUE
```


### Interpolation method: Inverse distance weigthing
```{r idw}
idw <- phylin::idw(
  values = earthquakes.sp.aea$mag,
  coords = earthquakes.sp.aea@coords,
  grid = grid,
  p = 2
)
idw.mag <- data.frame(grid, idw)
coordinates(idw.mag) <- ~ x + y
proj4string(idw.mag) <- aea
#gridded(idw.mag) <- TRUE


idw.mag.wgs84 <-
  sp::spTransform(idw.mag, "+proj=longlat +ellps=WGS84 +no_defs") %>% as.data.frame()
colnames(idw.mag.wgs84) <- c("mag", "lon", "lat")
```

Plot the interpolation result
```{r heat}
map3 <- ggplot() +
  # add the raster with coloring (CPU intensive!!!)
  geom_tile(data = idw.mag.wgs84, aes(lon, lat, fill = mag)) +
  geom_path(
    data = plates.sp,
    aes(long, lat, group = group),
    lwd = 2,
    color = colorblind[7]
  ) +
  borders(fill = alpha("grey", .25)) +
  scale_fill_scico() + # format legend labels
  coord_cartesian(
    xlim = range(earthquakes$longitude),
    ylim = range(earthquakes$latitude),
    expand = FALSE
  ) +
  labs(title = "Seismicity of western British Columbia",
       subtitle = "Data source: ANSS Comprehensive Earthquake Catalog",
       caption = "equal-area projection") +
  theme_bw()
map3
```

```{r, echo=FALSE, eval=FALSE, include=TRUE}
ggsave(map3, filename = "Data/DEM/my_heatmap.pdf", width = 11, height = 8.5)
```


## Useful resources

- https://erinbecker.github.io/r-raster-vector-geospatial/02-raster-plot/index.html
- https://www.earthdatascience.org/courses/earth-analytics/lidar-raster-data-r/crop-raster-data-in-r/
- Spatial interpolation with R: https://rspatial.org/raster/analysis/4-interpolation.html
- https://jjvhab.github.io/swathR/


```{r gc, include=FALSE}
gc(verbose = FALSE)
```


[previous course: Geochronology](R_turorial_4.html)
